---
title: "Shock and awe with R"
output:
  html_document:
    toc: yes
    toc_float: yes
  html_notebook:
    toc: yes
    toc_float: yes
---
[<<BACK](https://remi-daigle.github.io/2017-CHONe-Data/)

(Remi/David)
```{r,load packages,message=FALSE}
require(tidyverse)
require(sf)
require(robis)
require(marmap)
require(raster)
# require()
```

#Plot your study site
## Define site map limits
Let's start by defining a bounding box for your study site:

```{r, bounding box}
#define the corners
latmax <- 46.25
latmin <- 45.5
lonmax <- -61.25
lonmin <- -62.25

#create a matrix
bbmat <- cbind(
            c(lonmin,lonmax,lonmax,lonmin,lonmin),
            c(latmin,latmin,latmax,latmax,latmin)
        )
# make the matrix a 'simple features' polygon
bbsf <- st_polygon(list(bbmat))
# and let's give it information about the projection:
bbsfc <- st_sfc(bbsf,crs="+proj=longlat +datum=WGS84")
# finally, let's make it a simple features data.frame
bb <- st_sf(name="Study Site",geometry=bbsfc)
str(bb)

# plot it!
ggplot(bb)+
    geom_sf()
```

Riveting! 

##Add a basemap!

There are many ways to add a basemap in R

### marmap package

```{r,marmap}
bathymetry <- getNOAA.bathy(lonmin,lonmax,latmin,latmax,resolution=1,keep=TRUE)
plot.bathy(bathymetry,image=T,drawlabels=TRUE)

blues <- colorRampPalette(c("darkblue", "lightblue1"))
greys <- colorRampPalette(c(grey(0.4),grey(0.99)))

plot.bathy(bathymetry,
           image = TRUE,
           land = TRUE,
           n=0,
           bpal = list(c(0, max(bathymetry), greys(100)),
                       c(min(bathymetry), 0, blues(100))))

# this is a 'bathy' object and you could plot it as is with the tools provided in marmap, but to keep things 'simple' I'm going to convert to sf. There is no direct method yet, so... don't judge me!
# convert from bathy to raster to 'sp' polygon to simple features
# bathypoly <- marmap::as.raster(bathymetry) %>% 
    # rasterToPolygons() %>% st_as_sf
bathyras <- fortify.bathy(bathymetry)

bathyras$z[bathyras$z>0] <- NA

ggplot() +
    geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
    geom_sf(data=bb,fill="transparent",colour='black')
```


```{r}
Canada <- getData('GADM', country="CAN", level=1) %>% 
    st_as_sf() %>% 
    filter(NAME_1=="Nova Scotia"|
               NAME_1=="Prince Edward Island"|
               NAME_1=="New Brunswick")
ggplot() +
    geom_sf(data=Canada,aes(fill=NAME_1))

x=0.2
ggplot() +
    geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
    geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
    geom_sf(data=Canada,fill='grey40')+
    coord_sf(xlim = c(lonmin-x,lonmax+x),
             ylim = c(latmin-x,latmax+x),
             expand = F) 

```


#Get data from OBIS
 
OBIS is the [Ocean Biogeographic Information System](http://www.iobis.org/) and their vision is: "To be the most comprehensive gateway to the world’s ocean biodiversity and biogeographic data and information required to address pressing coastal and world ocean concerns."
 
 We can get access to their HUGE database through the excellent `robis` package by [ropensci](https://ropensci.org/) (unsurprisingly, they also have other great open science R packages)
 
 
```{r,eval=FALSE}
st_as_text(bb$geometry)
OBIS <- occurrence(geometry=st_as_text(bb$geometry))
write.csv(OBIS,'obis.csv',row.names = FALSE)
```

```{r}
OBIS <- read.csv("obis.csv")
OBIS <- st_as_sf(OBIS,
                 coords = c("decimalLongitude", "decimalLatitude"),
                 crs="+proj=longlat +datum=WGS84",
                 remove=FALSE) %>% 
    filter(!is.na(species))

ggplot() +
    geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
    geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
    geom_sf(data=Canada,fill='grey40')+
    geom_point(data=OBIS,aes(colour=phylum,x=decimalLongitude,y=decimalLatitude,size=2))+
    coord_sf(xlim = c(lonmin-x,lonmax+x),
             ylim = c(latmin-x,latmax+x),
             expand = F)
```
```{r}
ggplot(OBIS) +
    geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
    geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
    geom_sf(data=Canada,fill='grey40')+
    geom_point(data=OBIS,aes(x=decimalLongitude,y=decimalLatitude))+
    facet_wrap(~phylum)+
    coord_sf(xlim = c(lonmin-x,lonmax+x),
             ylim = c(latmin-x,latmax+x),
             expand = F)
```

[<<BACK](https://remi-daigle.github.io/2017-CHONe-Data/)
