---
title: "Shock and awe with R"
output:
  html_document:
    toc: yes
    toc_float: yes
  html_notebook:
    toc: yes
    toc_float: yes
---
[<<BACK](https://remi-daigle.github.io/2017-CHONe-Data/)

(Remi/David)

<!--

rm(list=ls())
setwd('/Users/davidbeauchesne/dropbox/phd/misc/2017-chone-data/2017-chone-data/')
library(rmarkdown)
render(input = 'shocknawe.rmd', 'html_document')

-->



<!-- ## Structure of document

Get people acquainted with existing possibilities in R from environmental point of view and oceanographic point of view

Structure for shock and awe portion 1.
Ecologically describe a study area from available open source data

1. Define study area
2.

- SDM
- Species interactions
- Environmental data
- Go look at ropensci
- wordcloud2

4. How about it's phylogeny?

    install.package('brranching')


Now that we have a pretty thorough description of the species, what else could we find out. How about known occurrence in our study area


How about fisheries activity?
install.package('rfisheries') -->

<!--
-->

```{r, message=FALSE}
library(sf)
library(marmap)
library(raster)
library(tidyverse)
library(ggplot2)
```

# IMAGINE!

Let's imagine that you are interested in a species in a given area and wish to know as much as possible about it. But, you can't go out in the field because funding is running short. What you do have, however, is a certain knowledge of the tools that are available to you and you are wondering how far they can take you. Let's find out.

## Defining species and area of interest

```{r, params}
    # Let's select Atlantic cod as our species of interest
    sp <- 'Gadus morhua'

    # And the gulf of St. Lawrence as our study area
    #define the corners
    latmax <- 52.01312
    latmin <- 45.52399
    lonmax <- -55.73636
    lonmin <- -71.06333
    #
    # latmax <- 46.25
    # latmin <- 45.5
    # lonmax <- -61.25
    # lonmin <- -62.25
```

## Describe your species

### Fishbase <!-- and Reol? -->

We'll start with a description of the species. First, let's see what [fishbase](http://www.fishbase.ca/) has to offer. This online data repository, along with [sealifebase](http://www.sealifebase.org/), contains a lot of information on marine and aquatic species all over the world and is accessible through the `package [rfishbase](https://github.com/ropensci/rfishbase)`

<!-- Reol, install.packages('reol') -->

```{r, ecology, eval = TRUE}
    # Let's have a look at the information available our species ecology
    ecol <- rfishbase::ecology(sp)
    ecol <- cbind(colnames(ecol), t(ecol))
    rownames(ecol) <- NULL
    ecol <- ecol[ecol[,2] != 0, ] # remove 0
    ecol <- ecol[!is.na(ecol[,2]), ] # remove NAs

    knitr::kable(ecol, col.names = c('Descriptors', 'Attributes'))
```

### Taxize

We can also get it's taxonomy rather easily. The package [taxize](https://ropensci.org/tutorials/taxize_tutorial.html) allows you to extract and validate, among other things, the taxonomy of millions of species by accessing an important number of online databases accessible through their Application program interface (API).

- Encyclopedia of Life (EOL) eol
- axonomic Name Resolution Service tnrs
- Integrated Taxonomic Information Service (ITIS) itis
- Global Names Resolver (from EOL/GBIF) gnr
- Global Names Index (from EOL/GBIF)
- IUCN Red List
- Tropicos (from Missouri Botanical Garden)
- Theplantlist.org
- Catalogue of Life
- National Center for Biotechnology Information
- CANADENSYS Vascan name search API
- International Plant Names Index (IPNI)
- World Register of Marine Species (WoRMS)
- Barcode of Life Data Systems (BOLD)
- Pan-European Species directories Infrastructure (PESI)
- Mycobank
- National Biodiversity Network (UK)
- Index Fungorum
- EU BON
- Index of Names (ION)
- Open Tree of Life (TOL)
- World Register of Marine Species (WoRMS)
- NatureServe

```{r, taxonomy, eval = TRUE}
    # Start by exporting the taxonomy of our species
        taxo <- taxize::classification(sp, db = 'worms', verbose = FALSE)
        taxo[[1]]

    # We can also extract the common or scientific names using sci2comm() & comm2sci(), respectively.
        taxize::sci2comm(sp, db = 'itis')

    # Or find out whether there are other names under which the species is known
        taxize::synonyms(sp, db = 'itis')

    # Another really interesting feature is to extract all known species at a given taxonomic scale.
    # Let's take the genus level and see the first 20 species that are part of that genus on the itis database
        taxize::children(strsplit(sp, ' ')[[1]][1], db = 'itis')[[1]]
```

### GloBI

How about extracting all known preys and predators of the species of interest? The [Global Biotic Interactions](http://globalbioticinteractions.org) web platform contains thousands of empirical binary interactions for multiple types of interactions, all over the world, and is easily accessible using the package [rglobi](https://github.com/ropensci/rglobi).

```{r, trophic, eval = TRUE}
    # There are multiple types of interactions available on GloBI
        rglobi::get_interaction_types()[,1:3]

    # For now let's focus on predator-prey interactions
        prey <- rglobi::get_prey_of(sp)$target_taxon_name
        pred <- rglobi::get_predators_of(sp)$target_taxon_name
        prey
        pred
```

<!-- Add d3 object or igraph object? -->

## Make your search spatially explicit

### Get data from OBIS

OBIS is the [Ocean Biogeographic Information System](http://www.iobis.org/) and their vision is: "To be the most comprehensive gateway to the world’s ocean biodiversity and biogeographic data and information required to address pressing coastal and world ocean concerns."

 We can get access to their HUGE database through the excellent `robis` package by [ropensci](https://ropensci.org/) (unsurprisingly, they also have other great open science R packages, from which this whole section is highly inspired)


```{r, OBIS data, eval=FALSE}
    sf::st_as_text(bb$geometry)
    OBIS <- robis::occurrence(scientificname = sp, geometry=st_as_text(bb$geometry), startdate = as.Date("2010-01-01"), enddate = as.Date("2017-01-01"), fields = c("species", "yearcollected","decimalLongitude", "decimalLatitude"))

    # remove duplicates
        OBIS <- unique(OBIS)

    write.csv(OBIS,'obis.csv',row.names = FALSE)
```

### Visualize data

We can easily visualize the data...

```{r, OBISPlot, eval = TRUE}
OBIS <- read.csv("obis.csv")
plot(OBIS[,c("decimalLongitude", "decimalLatitude")])
```

... but the aesthetics could be much better!

## Plot your study site
### Define site map limits
Let's start by defining a bounding box for your study site:

```{r, bounding box, eval = TRUE}
#create a matrix
bbmat <- cbind(
            c(lonmin,lonmax,lonmax,lonmin,lonmin),
            c(latmin,latmin,latmax,latmax,latmin)
        )
# make the matrix a 'simple features' polygon
bbsf <- st_polygon(list(bbmat))
# and let's give it information about the projection:
bbsfc <- st_sfc(bbsf,crs="+proj=longlat +datum=WGS84")
# finally, let's make it a simple features data.frame
bb <- st_sf(name="Study Site",geometry=bbsfc)
str(bb)

# plot it!
ggplot(bb) + geom_sf()
```

### Add a basemap!

There are many ways to add a basemap in R

#### marmap package

```{r,marmap, eval = TRUE}
bathymetry <- getNOAA.bathy(lonmin,lonmax,latmin,latmax,resolution=1,keep=TRUE)
plot.bathy(bathymetry,image=T,drawlabels=TRUE)

blues <- colorRampPalette(c("darkblue", "lightblue1"))
greys <- colorRampPalette(c(grey(0.4),grey(0.99)))

plot.bathy(bathymetry,
           image = TRUE,
           land = TRUE,
           n=0,
           bpal = list(c(0, max(bathymetry), greys(100)),
                       c(min(bathymetry), 0, blues(100))))

# this is a 'bathy' object and you could plot it as is with the tools provided in marmap, but to keep things 'simple' I'm going to convert to sf. There is no direct method yet, so... don't judge me!
# convert from bathy to raster to 'sp' polygon to simple features
# bathypoly <- marmap::as.raster(bathymetry) %>%
    # rasterToPolygons() %>% st_as_sf
bathyras <- fortify.bathy(bathymetry)

bathyras$z[bathyras$z>0] <- NA

x=0.2
ggplot() +
    geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
    geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
    # geom_sf(data=Canada,fill='grey40')+
    coord_sf(xlim = c(lonmin-x,lonmax+x),
             ylim = c(latmin-x,latmax+x),
             expand = F)
```

## Overlay your species occurrences over the base map

You can now overlay your species occurrences on your base map!

```{r, eval = TRUE}
OBIS <- st_as_sf(OBIS,
                 coords = c("decimalLongitude", "decimalLatitude"),
                 crs="+proj=longlat +datum=WGS84",
                 remove=FALSE) %>%
    filter(!is.na(species))


    ggplot(OBIS) +
        geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
        geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
        # geom_sf(data=Canada,fill='grey40')+
        geom_point(data=OBIS,aes(x=decimalLongitude,y=decimalLatitude))+
        facet_wrap(~species)+
        coord_sf(xlim = c(lonmin-x,lonmax+x),
                 ylim = c(latmin-x,latmax+x),
                 expand = F)
```

## Environmental variables

There are also multiple resources to access environmental data. As seen before, we can easily access bathymetry data using the marmap package. Other environmental data can also be accessed using other packages like `raster` for terrestrial climatic data and `rnoaa`

### Extract sea surface temperature (SST)

```{r, NOAA, eval = TRUE}
    # Bear with us, for simplicity's sake we will only take one year of data for one month, knowing very well that this does not make sense ecologically!
    # Eventually use apply() to load multiple datasets and create rasterStack
        # An exemple with sea ice
            # urls <- seaiceeurls(mo='Apr', pole='N')[1:10]
            # out <- lapply(urls, seaice)
            # names(out) <- seq(1979,1988,1)
            # df <- ldply(out)
            # library('ggplot2')
            # ggplot(df, aes(long, lat, group=group)) +
            # geom_polygon(fill="steelblue") +
            # theme_ice() +
            # facet_wrap(~ .id)

        sst.nc <- rnoaa::ersst(2010, 8, overwrite = TRUE)
        sst <- raster(sst.nc$filename, varname = 'sst')
        sst@data@values <- values(sst) # make sure values are in the right data slot

        # never do that...
        sst@extent@xmin <- sst@extent@xmin - 361
        sst@extent@xmax <- sst@extent@xmax - 361

    # Transform rasterLayer as a data.frame
        sst.spdf <- as(sst, "SpatialPixelsDataFrame")
        sst.df <- as.data.frame(sst.spdf)
        colnames(sst.df) <- c('sst','x','y')

    # Plot newly acquired data!
        ggplot(sst.df, aes(x=x, y=y)) +
            geom_tile(aes(x=x,y=y,fill=sst)) +
            coord_equal()

```

## Species distribution models

Now what we can do once we have species occurrence data and environmental variables are species distribution models (SDMs). There are multiple ways and packages to perform SDMs and this [vignette](https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf) provides an extensive overview of the different methods available. For this workshop, we will be using the `package biomod2`, which implements most of the methods presented in the vignette. There is also a thorough explanation of our to use biomod2 in another very interesting [vignette](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/pkg/biomod2/inst/doc/Simple_species_modelling.pdf?root=biomod)


### Formating data

The first thing to do is to format our data to be able to work with biomod2 functions, which work with BIOMOD.formated.data.PA class objects. This entails stacking all of our environmental data together to perform analyses.

```{r, envCov}

    # Create raster stack with environmental variables
    # raster have to be of similar extent and resolution
        # Load bathy data from marmap as raster
            bathypoly <- marmap::as.raster(bathymetry)
            bathypoly@data@values[bathypoly@data@values > 0] <- NA

        # Resample sst to make it same extent as bathypoly
            # create empty raster with appropriate resolution and extent
                sst2 <- raster(resolution = res(bathypoly), ext = extent(bathypoly), crs=bathypoly@crs)
            # Beware, this may not be good practice because we are changing the resolution of the data to a finer resolution!
                sst2 <- resample(sst, sst2, method='bilinear') # extract data from sst

        # Create raster stack
            envCov <- raster::stack(bathypoly, sst2)

        plot(envCov)
```

You can then create your biomod2 data class.

```{r, SDM data}
    OBIS <- read.csv("obis.csv")
    SDMdata <- biomod2::BIOMOD_FormatingData(resp.var = rep(1,nrow(OBIS)),
                                            expl.var = envCov,
                                            resp.xy = OBIS[, c("decimalLongitude","decimalLatitude")],
                                            resp.name = sp,
                                            PA.nb.rep = 2)
    SDMdata
    plot(SDMdata)
```

You can then use this object with the biomod2 functions are generate a distribution model for your species of interest.

```{r, SDM model}
    # Basic options for modelling
        SDMoption <- biomod2::BIOMOD_ModelingOptions()

    # SDM model
        SDM <- biomod2::BIOMOD_Modeling(data = SDMdata,
                                        models = "MAXENT.Tsuruoka",
                                        model.options = SDMoption)

    # Print model summary
        SDM

    # print the dimnames of this object to understand the structure of the returned objects
        dimnames(SDM)

    # get all models evaluation.
    # These give you an evaluation of the performance of your model.
        biomod2::get_evaluations(SDM)

    # Finally, once the models are calibrated and evaluated, you can project your species spatial distribution...
        SDMproj <- biomod2::BIOMOD_Projection(modeling.output = SDM,
                                              new.env = envCov,
                                              proj.name = 'Awesome!',
                                              selected.models = 'all',
                                              binary.meth = 'TSS',
                                              compress = 'xz',
                                              clamping.mask = F,
                                              output.format = '.grd')
    # ... and visualize them!                
        plot(SDMproj)
```

Riveting! We starting with a basic idea and ended up with a species description, an idea of the distribution of certain environmental parameters in our area of interest (could be better, we know!) and the potential spatial distribution of our species!

...but should we stop there?

The reality is, there is so much more going on in our area of interest than the spatial distribution of a single species. Let's see what other insights we an get from available open source resources.



<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->
<!-- ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->



## Community level analyses

One of the usual challenges in ecology is to perform analyses at the community scale often due to the difficulty of sampling multiple species in a single system. The reality is that our species of interest is very likely to interact and be impacted by other species.

### Extract all occurrence known occurrence data in area of interest

OBIS once again gives us the opportunity to access multiple species occurrences. Let's go back to OBIS and extract every available data in our system.

```{r, multi species OBIS, eval = FALSE}
    # Load all known species occurrences from our area of interest
        multiOBIS <- robis::occurrence(geometry=st_as_text(bb$geometry), startdate = as.Date("2010-01-01"), enddate = as.Date("2017-01-01"), fields = c("species", "yearcollected","decimalLongitude", "decimalLatitude"))
        multiOBIS <- multiOBIS[!is.na(multiOBIS[,'species']), ] # remove species == NA

    # Remove duplicates
        multiOBIS <- unique(multiOBIS)

    # Select species for which there is at least n observations
        n = 50
        spOK <- names(which(table(multiOBIS[,'species']) >= n)) # id of species with at least n observations
        multiOBIS <- multiOBIS[multiOBIS[, 'species'] %in% spOK, ] # Subset of multiOBIS data with species with at least n observations
        multiOBIS[,'species'] <- as.factor(multiOBIS[,'species'])
        multiOBIS[,'yearcollected'] <- as.factor(multiOBIS[,'yearcollected'])
        write.csv(multiOBIS,'multiOBIS.csv',row.names = FALSE)
```

```{r, multiOBISPlot, eval = T}

    # Read occurrence data file
        multiOBIS <- read.csv("multiOBIS.csv")

    # Structure of object
        str(multiOBIS)

    # Visualize species list
        spList <- unique(multiOBIS[,'species'])
        length(spList)
        spList

    # Plot
        multiOBIS <- sf::st_as_sf(multiOBIS,
                         coords = c("decimalLongitude", "decimalLatitude"),
                         crs="+proj=longlat +datum=WGS84",
                         remove=FALSE) %>%
            filter(!is.na(species))

        ggplot(multiOBIS) +
            geom_raster(data=bathyras,aes(x=x,y=y,fill=z))+
            geom_sf(data=bb,fill="transparent",colour='yellow',size=2)+
            geom_point(data=multiOBIS,aes(x=decimalLongitude,y=decimalLatitude))+
            coord_sf(xlim = c(lonmin-x,lonmax+x),
                     ylim = c(latmin-x,latmax+x),
                     expand = F)
```

### Trophic network

```{r, interaction matrix, eval = FALSE}
    nsp <- length(spList)
    foodWeb <- matrix(ncol = nsp, nrow = nsp, data = 0, dimnames = list(spList, spList))

    pb <- txtProgressBar(min = 0,max = nsp, style = 3)
    for(i in 1:nsp) {
        for(j in 1:nsp) {
            inter <- rglobi::get_interactions_by_taxa(sourcetaxon = spList[i], targettaxon = spList[j], interactiontype = 'eats')
            if(nrow(inter) > 0) foodWeb[j,i] <- 1
        }
    setTxtProgressBar(pb, i)
    }
    close(pb)

    saveRDS(foodWeb, file = './foodWeb.rds')
```

We can now visualize the network using the package iGraph

```{r, foodWeb iGraph}
    nsp <- length(spList)
    foodWeb <- readRDS('./foodWeb.rds')

    netw <-  igraph::graph_from_adjacency_matrix(t(foodWeb))

    vec_col <- spList
    vec_col[vec_col != sp] <- "#11bdec"
    vec_col[vec_col == sp] <- "#cc154b"


    plot(netw,
      vertex.color = vec_col,
      vertex.frame.color = "transparent",
      vertex.label.color = "black",
      edge.arrow.size = .5
      )
```

... Not very nice visually when there are too many species. We can however do this better (and more fun!) using the package networkD3

```{r, d3network}

netwD3 <- networkD3::igraph_to_networkD3(netw, group = vec_col, what = 'both')
nodeSize <- spList
nodeSize[nodeSize != sp] <- 2
nodeSize[nodeSize == sp] <- 4
netwD3$nodes <- cbind(netwD3$nodes, nodeSize)

networkD3::forceNetwork(Links = netwD3$links,
                        Nodes = netwD3$nodes,
                        Source = 'source',
                        Target = 'target',
                        NodeID = 'name',
                        Group = 'group',
                        zoom = TRUE,
                        linkDistance = 50,
                        fontSize = 12,
                        opacity = 0.9,
                        charge = -20,
                        # colourScale = networkD3::JS('d3.scale.ordinal().range(vec_col)'),
                        Nodesize = 'nodeSize')

```

Neat huh?!?

### Co-distribution

We could also evaluate how species are co-distributed using an analysis analog to SDMs, but at the communicty scale and using a hierarchical Bayesian approach, Hierarchical Modeling of Species Communities (HMSC; **ref**) currently being implemented in R in the package `HMSC` (**ref**). As with SDM using the `biomod2` package, the first step is to format the data into a HMSC class object.

#### Formatting the data

##### Create regular grid over study area

In order to create common observations among individual observations, we will aggregate all data within regular grid cells, so that cells will become planning units on which we base co-distribution of species. We will also create a rather coarse sized grid in order to make processing speed more efficient. Of course there would be more sophisticated ways to do this, but for simplicity's sake we will perform a simple intersect between the grid and occurrence data.

```{r, gridFunction, eval = FALSE}
    # First create a function to generate hexagonal grids
        hexGrid <- function(size, shp_over) {
            library(rgeos)
            library(sp)
            shp_over2 <- gBuffer(shp_over, width = size) # buffer so that whole surface of shapefile is covered by a cell
            spat_grid <- spsample(shp_over2,type="hexagonal",cellsize=size, offset=c(0,0)) # creating points for a hex grid
            spat_grid <- HexPoints2SpatialPolygons(spat_grid) # creating polygons
            proj4string(spat_grid) <- proj4string(shp_over) # set projection
            spat_grid <- spat_grid[shp_over, ] # clipping grid to retain only cells intersecting shapefile

            id <- paste('ID',seq(1:length(spat_grid)), sep = '')
            row.names(spat_grid) <- id
            spat_grid <- SpatialPolygonsDataFrame(Sr = spat_grid, data = data.frame(ID = id, row.names = id))

            return(spat_grid)
        }

    # Load shapefile over which grid is made
        egsl <- rgdal::readOGR(dsn = "./", layer = "egsl")

    # Create grid
        egsl_grid <- hexGrid(size = 10000, shp_over = egsl)
        rgdal::writeOGR(egsl_grid, dsn = './', layer = 'egsl_grid', driver="ESRI Shapefile", overwrite_layer=TRUE)

    # Plot grid
        plot(egsl_grid)
```

```{r, plotGrid, eval = TRUE, message = FALSE}
    # Plot grid
        egsl_grid <- rgdal::readOGR(dsn = "./", layer = "egsl_grid")
        proj <- "+proj=longlat +datum=WGS84"
        egsl_grid <- sp::spTransform(egsl_grid, CRSobj = CRS(proj))
        plot(egsl_grid)
```

##### Intersect occurrences and environmental data with grid

```{r, occGrid, eval = TRUE}
    # reload multiOBIS data and transform into sp points
        multiOBIS <- read.csv("multiOBIS.csv")
        proj <- "+proj=longlat +datum=WGS84"
        multiOBISpt <- sp::SpatialPointsDataFrame(coords = multiOBIS[, c("decimalLongitude", "decimalLatitude")], data = multiOBIS, proj4string = CRS(proj), match.ID = TRUE)

    # Overlay occurrences on grid
        occGrid <- sp::over(multiOBISpt, egsl_grid)
        multiOBIS <- cbind(multiOBIS, occGrid)
        multiOBIS_wide <- reshape2::dcast(multiOBIS, formula = ID ~ species, value.var = 'species', fun.aggregate = length)

        for(i in 2:ncol(multiOBIS_wide)) {
            for(j in 1:nrow(multiOBIS_wide)) {
                if(multiOBIS_wide[j,i] > 1) multiOBIS_wide[j,i] <- 1
            }
        }

        toAdd <- egsl_grid@data[,'ID'][!egsl_grid@data[,'ID'] %in% multiOBIS[,'ID']]
        matSup <- matrix(nrow = length(toAdd), ncol = ncol(multiOBIS_wide), dimnames = list(c(), colnames(multiOBIS_wide)), data = 0)
        matSup[,'ID'] <- toAdd
        multiOBIS_wide <- rbind(multiOBIS_wide, matSup)
        multiOBIS_wide <- multiOBIS_wide[-which(is.na(multiOBIS_wide[,'ID']) == T), ]
        egsl_grid@data <- merge(egsl_grid@data, multiOBIS_wide, by = 'ID')
        spAll <- colnames(egsl_grid@data)[2:ncol(egsl_grid@data)]
        spID <- seq(1:length(spAll))
        colnames(egsl_grid@data)[2:ncol(egsl_grid@data)] <- spID

    # Extract environmental values from rasters
        egsl_grid@data <- cbind(egsl_grid@data, raster::extract(bathypoly, egsl_grid, fun = mean, na.rm = T))
        colnames(egsl_grid@data)[ncol(egsl_grid@data)] <- 'bathy'
        egsl_grid@data <- cbind(egsl_grid@data, raster::extract(sst2, egsl_grid, fun = mean, na.rm = T))
        colnames(egsl_grid@data)[ncol(egsl_grid@data)] <- 'sst'

    # HMSC data
        studyGrid <- egsl_grid@data

        envCov <-  c('bathy','sst')

    # Remove environmental NAs
        NAs <- studyGrid[, envCov] %>%
                lapply(X = ., FUN = function(x) which(is.na(x))) %>%
                unlist(.) %>%
                unique(.)
        NAsID <- studyGrid[NAs,'ID']

        studyGrid <- studyGrid[-NAs, ]
```

##### Create HMSC class data

```{r, HMSC data}
    # ---------------------------------
    # Y: sample units by species matrix
    # ---------------------------------
        # Extracting only presence/absence data
            Station <- studyGrid[, 'ID']
            Y <- studyGrid[, as.character(spID)]
            rownames(Y) <- Station

            for(i in 1:ncol(Y)) {
                Y[,i] <- as.numeric(as.character(Y[,i]))
            }

    # -----------------------------------------
    # Pi: sample units by random effects matrix
    # -----------------------------------------
        # Create a dataframe for random effects, columns have to be factors
            Pi <- data.frame(sampling_unit = as.factor(Station))

    # ----------------------------------------------------
    # X: sampling units by environmental covariates matrix
    # ----------------------------------------------------
        # Create a matrix for the values of environmental covariates at each sampling unit location
            X <- matrix(ncol = length(envCov),
                        nrow = nrow(studyGrid))
            X <- studyGrid[, envCov]
                    rownames(X) <- Station

    # ----------------------------
    # as.HMSCdata for HMSC package
    # ----------------------------
        # Creating HMSC dataset for analyses
            HMSCdata <- HMSC::as.HMSCdata(Y = Y, X = X, Random = Pi, interceptX = T, scaleX = T)
            str(HMSCdata)
```

##### HMSC analysis

We are now ready to perform the analyses!

```{r, HMSC model, eval = FALSE}
    model <- HMSC::hmsc(HMSCdata,
                        family = "probit",
                        niter = 10000,
                        nburn = 100,
                        thin = 10)
    saveRDS(model, file = './HMSCmodel.rds')
```

##### HMSC analysis diagnostics

```{r, MCMC sampling}
model <- readRDS('./HMSCmodel.rds')
# =========================================
# 3. Producing MCMC trace and density plots
# =========================================
        mixingMeansParamX <- coda::as.mcmc(model, parameters = "meansParamX")

    # Trace and density plots to visually diagnose mcmc chains
    # Another way to check for convergence is to use diagnostic tests such as Geweke's convergence diagnostic (geweke.diag function in coda) and the Gelman and Rubin's convergence diagnostic (gelman.diag function in coda).
        paramModel <- colnames(mixingMeansParamX)
        nParam <- length(paramModel)

        par(mfrow = c(nParam, 2), mar = rep(2, 4))
        for(i in 1:ncol(mixingMeansParamX)) {
          coda::traceplot(mixingMeansParamX[,i], col = "blue", main = paste('Trace of ', paramModel[i]))
          coda::densplot(mixingMeansParamX[,i], col = "orange", main = paste('Density of ', paramModel[i]))
        }
```

```{r, posterior summaries}
# ================================
# 4. Producing posterior summaries
# ================================
    # Average
        average <- apply(model$results$estimation$paramX, 1:2, mean)
    # 95% confidence intervals
        CI.025 <- apply(model$results$estimation$paramX, 1:2, quantile, probs = 0.025)
        CI.975 <- apply(model$results$estimation$paramX, 1:2, quantile, probs = 0.975)

    # Summary table
        paramXCITable <- cbind(unlist(as.data.frame(average)),
                             unlist(as.data.frame(CI.025)),
                             unlist(as.data.frame(CI.975)))
        colnames(paramXCITable) <- c("average", "lowerCI", "upperCI")
        rownames(paramXCITable) <- paste(rep(colnames(average), each = nrow(average)), "_", rep(rownames(average), ncol(average)), sep="")

    # Credible intervals
        paramXCITable_Full <- paramXCITable

        beg <- seq(1,nrow(paramXCITable_Full), by = nsp)
        end <- seq(nsp,nrow(paramXCITable_Full), by = nsp)
        sign <- abs(as.numeric(paramXCITable_Full[, 'lowerCI'] <= 0 & paramXCITable_Full[, 'upperCI'] >=0) - 3)

    # Export figure
        par(mfrow = c((length(beg)+1),1))
        for(i in 1:length(beg)) {
            paramXCITable <- paramXCITable_Full[beg[i]:end[i], ]
            cols <- sign[beg[i]:end[i]]
            par(mar=c(1,2,1,1))
            plot(0, 0, xlim = c(1, nrow(paramXCITable)), ylim = round(range(paramXCITable)), type = "n", xlab = "", ylab = "", main=paste(colnames(mixingMeansParamX)[i]), xaxt="n", bty = 'n')
            axis(1,1:nsp,las=2, labels = rep('',nsp))
            abline(h = 0,col = 'grey')
            arrows(x0 = 1:nrow(paramXCITable), x1 = 1:nrow(paramXCITable), y0 = paramXCITable[, 2], y1 = paramXCITable[, 3], code = 3, angle = 90, length = 0.05, col = cols)
            points(1:nrow(paramXCITable), paramXCITable[,1], pch = 15, cex = 1, col = cols)
        }
        mtext(text = spAll, side = 1, line = 1, outer = FALSE, at = 1:nsp, col = 1, las = 2, cex = 0.4)
```

```{r, association networks}
# =======================
# 6. Association networks
# =======================
    # Extract all estimated associatin matrix
        assoMat <- HMSC::corRandomEff(model)

    # Average
        plotMean <- apply(assoMat[, , , 1], 1:2, mean)
        colnames(plotMean) <- rownames(plotMean) <- spAll

    # Build matrix of colours for chordDiagram
        plotDrawCol <- matrix(NA, nrow = nrow(plotMean), ncol = ncol(plotMean))
        plotDrawCol[which(plotMean > 0.4, arr.ind=TRUE)]<-"red"
        plotDrawCol[which(plotMean < -0.4, arr.ind=TRUE)]<-"blue"
    # Build matrix of "significance" for corrplot
        plotDraw <- plotDrawCol
        plotDraw[which(!is.na(plotDraw), arr.ind = TRUE)] <- 0
        plotDraw[which(is.na(plotDraw), arr.ind = TRUE)] <- 1
        plotDraw <- matrix(as.numeric(plotDraw), nrow = nrow(plotMean), ncol = ncol(plotMean))

    # plots
        # Matrix plot
        Colour <- colorRampPalette(c("red", "white", "blue"))(200)
        corrplot::corrplot(plotMean, method = "color", col = Colour, type = "lower", diag = FALSE, p.mat = plotDraw, tl.srt = 65, tl.cex = 0.4, pch.cex = 0.3, tl.col = 'black')

        # Chord diagram
        circlize::chordDiagram(plotMean, symmetric = TRUE, annotationTrack = 'grid', grid.col = "grey", col = plotDrawCol, preAllocateTracks = 1)
        circlize::circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
          xlim = circlize::get.cell.meta.data("xlim")
          ylim = circlize::get.cell.meta.data("ylim")
          sector.name = circlize::get.cell.meta.data("sector.index")
          circlize::circos.text(mean(xlim), ylim[1] + .1, sector.name, facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex = 0.4)
        }, bg.border = NA)

```

```{r, predictive power}
    # ===============================================
    # 7. Computing the explanatory power of the model
    # ===============================================
        # Prevalence
            prevSp <- colSums(HMSCdata$Y)
        # Coefficient of multiple determination
            R2 <- HMSC::Rsquared(model, averageSp = FALSE)

        # Draw figure
            plot(prevSp, R2, xlab = "Prevalence", ylab = expression(R^2), cex = 0.8, pch=19, las=1, cex.lab = 1, main = 'Explanatory power of the model')
            abline(h = mean(R2, na.rm = T), col = "blue", lwd = 2)
```

```{r, AUC}
    # =============================================
    # 8. Generating predictions for validation data
    # =============================================
        source('./crossValid.r')
        modelAUC <- crossValidation(data = HMSCdata,  nCV = 2, validPct = 0.2, as.strata = FALSE)

        meanAUC <- colMeans(modelAUC)
        sdAUC <- apply(modelAUC, MARGIN = 2, FUN = sd)

        plot(prevSp, meanAUC, ylim = c(0,1), xlab = "Prevalence", ylab = 'AUC', cex = 0.8, pch=19, las=1, cex.lab = 1, main = 'Monte Carlo cross-validation with AUC of ROC curves')
        abline(h = 0.5, col = 'grey')
        arrows(x0 = prevSp, x1 = prevSp, y0 = (meanAUC - sdAUC), y1 = (meanAUC + sdAUC), code = 3, angle = 90, length = 0.05)
        points(prevSp, meanAUC, pch = 19, cex = 0.8)
```

```{r, predictions}
# Generate predictions
    HMSCpred <- predict(model)

# Replace NAs in grid
    data <- matrix(nrow = (length(NAs) + nrow(HMSCpred)), ncol = ncol(HMSCpred), data = 0)
    dimnames(data) = list(paste('ID',seq(1:nrow(data)), sep = ''), colnames(HMSCpred))

    for(i in 1:length(NAsID)) {
        data[which(rownames(data) == NAsID[i]), ] <- NA
    }

    for(i in 1:nrow(HMSCpred)) {
        data[which(rownames(data) == rownames(HMSCpred)[i]), ] <- HMSCpred[i, ]
    }

    HMSCpred <- data
    HMSCpred <- cbind(HMSCpred, rownames(HMSCpred))
    colnames(HMSCpred)[ncol(HMSCpred)] <- 'ID'
    # colnames(HMSCpred) <- spAll
```


```{r, plotPred, eval = TRUE, message = FALSE}
    # Plot grid
        egsl_grid <- rgdal::readOGR(dsn = "./", layer = "egsl_grid")
        proj <- "+proj=longlat +datum=WGS84"
        egsl_grid <- sp::spTransform(egsl_grid, CRSobj = CRS(proj))


        egsl_grid@data <- merge(egsl_grid@data, HMSCpred, by = 'ID')

        data <- egsl_grid@data[,31]
        rbPal <- colorRampPalette(c('#2f6eb9','#2aadba','#b45f5f')) # color palette
        cols <- rbPal(50)[as.numeric(cut(as.numeric(data), breaks = 50))]
        plot(egsl_grid, col = cols, border = cols)

```


[leaflet example](https://letir.github.io/Elections_fr-042017/)


[<<BACK](https://remi-daigle.github.io/2017-CHONe-Data/)
